An open and continuously updated fern tree of life (FTOL)

### 2022-01-17 ms/manuscript.Rmd
Update MS Rmd

```{r load-targets}
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_tree,
		plastome_outgroups,
		ppgi_taxonomy),
	store = here::here("ftol_cache")
)
```

```{r abstract-stats}
# Make tibble with one row per species in Sanger sampling
sanger_sampling <- tibble(
	species = rownames(sanger_alignment)
) %>%
	# Add higher-level taxonomy
	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, order), by = "genus"
	)

# Count number of fern species in Sanger data
n_sanger_ferns <-
	sanger_sampling %>% 
	# outgroup taxa have no family data
	filter(!is.na(family)) %>%
	pull(species) %>%
	n_distinct()
```

@@

The plastid genome has been the most widely sequenced genomic compartment in ferns by far, so we used plastid loci to build our tree.
Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome genes").
*trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*.
*rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication.

@@

To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of known sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019]
Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.
We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
We then filtered the reference sequences to a single representative longest sequence per genus (Sanger loci only).
Next, we aligned the sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" on the sequences downloaded from GenBank, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences to use as references that lacked annotations in the first round.
The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences.

@@

>>>>

```{r load-targets}
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_tree,
		plastome_metadata_renamed,
		ppgi_taxonomy),
	store = here::here("ftol_cache")
)
```

```{r abstract-stats}
# Make tibble of outgroup species
og_species <- 
	plastome_metadata_renamed %>%
	select(species, outgroup) %>%
	filter(outgroup == TRUE)

# Make tibble with one row per species in Sanger sampling
sanger_sampling <- tibble(
	species = rownames(sanger_alignment)
) %>%
	# Add higher-level taxonomy
	mutate(genus = str_split(species, "_") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, order), by = "genus"
	) %>% 
	# Add outgroup status
	left_join(og_species, by = "species") %>%
	mutate(outgroup = replace_na(outgroup, FALSE)) %>%
	verify(nrow(.) == nrow(sanger_alignment)) %>%
	verify(sum(outgroup) == nrow(og_species))

# Count number of fern species in Sanger data
n_sanger_ferns <-
	sanger_sampling %>% 
	filter(outgroup == FALSE) %>%
	n_distinct(.$species)
```

@@

The plastid genome has been the most widely sequenced genomic compartment in ferns by far, so we used plastid loci to build our tree.
Our sampling includes two major categories of sequence data: seven loci (*atpA*, *atpB*, *matK*, *rbcL*, *rps4*, *trnL*--*trnF*, and *rps4*--*trnS*) that have been frequently used for molecular systematics studies in ferns and are typically obtained by PCR and Sanger sequencing ("Sanger loci") and a much larger set of single-copy, coding genes typically obtained from next-generation sequencing of the plastome ("plastome genes").
*trnL*--*trnF* includes the *trnL* intron, the 3' *trnL* exon, and the spacer between *trnL* and *trnF*.
*rps4*--*trnS* includes the spacer between *rps4* and *trnS*.
The set of plastome genes is based on the list of 83 protein-coding genes of @Wei2017a, which was filtered to only genes that show no evidence of duplication (77 genes).

@@

To avoid missing sequences due to differences in annotation format, we used the "Reference_Blast_Extract.py" script of  superCRUNCH [@Portik2020] to extract target loci from GenBank FASTA files.
Briefly, this involves querying candidate FASTA files downloaded from GenBank with BLAST [@Altschul1997] against a reference database of full length, representative sequences [i.e., a "baited search"\; @Smith2009b; @Smith2019].
Those portions of the query that have a significant match in the reference are extracted and written to a new filtered FASTA file.

We constructed the superCRUNCH reference databases by first downloading fern sequences from GenBank, then extracting target gene sequences with a custom R script that parses the GenBank flatfile; this only works for properly annotated accessions.
We then filtered the sequences to a single representative longest sequence per genus.
Next, we aligned the filtered sequences with MAFFT [@Katoh2002] and removed poorly aligned regions with trimAl [@Capella-Gutierrez2009].
To maximize the size of the reference database for Sanger loci, we then ran "Reference_Blast_Extract.py" using these sequences as references, followed by filtering to the longest sequence per genus and alignment and cleaning as before; this retrieved additional sequences that lacked annotations in the first round.
The cleaned alignments were then used as references for superCRUNCH to obtain a maximally sampled set of fern sequences from sequences downloaded from GenBank.

@@

## Sequence selection and concatenation

A typical step in multilocus workflows is to concatenate loci across samples.
For phylogenetic studies aiming to produce one tip per species, this is often done by concatenating the longest sequence per locus within each species.
This is potentially problematic because accessions on GenBank may be misidentified and/or species may not be monophyletic.
Therefore, we concatenated accessions and selected one final set of concatenated loci per species as follows (here, we refer to "accession" to mean a single locus within a GenBank accession; GenBank accessions sometimes have multiple loci per accession, but we already split those out using superCRUNCH as described previously).
First, we constructed gene trees with FastTree [@Price2009; @Price2010] on default settings, and classified species as monophyletic if they were monophyletic in all gene trees including multiple accessions of that species.
We then concatenated loci for monophyletic species by selecting the longest accession per locus within each species.
Next, for the remaining species, we concatenated loci across accessions if all accessions originated from the same voucher specimen; this could not be done in all cases as voucher information is not always annotated on GenBank.
Next, for the remaining species, we concatenated loci across accessions if all accessions for a given species originated from only one publication, on the assumption that the authors of the study intended to do so.
For the remaining species that could not be concatenated using any of the above methods, we selected one longest accession per locus and discarded the others.

We then selected the final set of concatenated loci for each species. 
We sought to maximize representation of *rbcL* as this is historically the most sequenced locus for ferns, and we expect that maximal sampling of one locus should improve phylogenetic analysis.
Our selection is based on the following criteria, in order of priority:

1. The sample (set of accessions joined by the methods described above) with the longest concatenated sequence length; sample includes *rbcL* and additional loci.
2. The sample with the longest *rbcL* sequence; sample includes only *rbcL* and no additional loci.
3. The sample with the longest concatenated sequence length; sample does not include *rbcL*.

The above steps were not needed for plastome data, as each plastome sequence on GenBank originates from a single voucher specimen.
For these, we selected one specimen per species with the longest combined sequence length across all plastome genes.

## Sequence alignment

For non-spacer regions, we aligned each locus separately with MAFFT.

Spacer regions are difficult to align across higher taxonomic levels within ferns (e.g., family or higher) as they include frequent indels; however, spacer regions are very useful for phylogenetic analysis at finer levels (e.g., within genera) where slower-evolving genes like *rbcL* may not provide enough resolution.
Therefore, we first aligned spacer regions for each family with MAFFT, then merged the subalignments using the MAFFT "--merge" option.
This retains the indels within each subalignment while aligning across subalignments.
For some small families with fewer than three species each that could be used for subalignments, these were added as "singletons" during the MAFFT "--merge".

Next, we removed poorly aligned regions from the alignments using trimAl.
We removed regions with >1% or >5% of sequences having gaps from coding regions or spacers, respectively.

## Phylogenetic analysis
